Modern BTree\\ techniques

### 2020-01-22 btree/presentation.tex
Add "more" section
Add notes and MDAM

\tikzset{rum-node/.style={draw,circle,fill=white,minimum width=2cm}}
\tikzset{rum-extra-node/.style={rectangle, draw, dashed, rounded corners, fill=white }}
\tikzset{btree-key/.style={minimum height=1cm, pattern=north west lines, draw}}
\tikzset{btree-pointer/.style={minimum height=1cm, draw}}
\tikzset{btree-hide/.style={draw opacity=0, line width=0, pattern=none}}
\tikzset{btree-empty/.style={btree-pointer, dashed}}
\tikzset{btree-leaf/.style={fill=greenGood}}
\tikzset{btree-branch/.style={fill=gray!20}}
\tikzset{btree-line/.style={line width=0.5pt}}
\tikzset{btree-path/.style={btree-pointer, minimum width=0.8cm, fill=red, opacity=0.5}}
\tikzset{>=latex}
\tikzset{btree-high-key/.style={btree-key, pattern=north west lines, pattern color=red, draw=red}}
\tikzset{btree-key-prefix/.style={btree-key, minimum height=0.2cm, fill=red, opacity=0.5}}
\tikzset{btree-key-compare/.style={btree-key, minimum height=0.6cm, fill=blue, opacity=0.5}}
\tikzset{arrow-pointer/.style={
        single arrow,
        minimum height=1.5cm,
        inner sep=3pt,
        line width=1pt,
        draw,
        color=gray,
        single arrow tip angle=45,
        single arrow head extend=0.1cm
    }
}
\tikzset{btree-indirection/.style={btree-key, minimum width=0.2cm, inner sep=0, pattern=none}}
\tikzset{btree-page-header/.style={btree-key, minimum width=0.5cm, inner sep=0, pattern=none}}
\tikzset{cpu-cache/.style={draw, minimum width=1cm, minimum height=1cm, text width=1cm, align=center, rounded corners}}
\tikzset{cpu-cache-hide/.style={cpu-cache, draw=none, color=white}}
\tikzset{btree-node/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners}}
\tikzset{btree-leaf-node/.style={btree-node, fill=greenGood}}
\tikzset{btree-branch-node/.style={btree-node, fill=gray!20}}
\tikzset{btree-partition1/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners, fill=red!20}}
\tikzset{btree-partition2/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners, fill=green!20}}
\tikzset{btree-partition3/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners, fill=blue!20}}
\tikzset{smallbtree-node/.style={btree-node, minimum width=0.4cm, minimum height=0.4cm, rounded corners=0.1cm}}
\tikzset{box/.style={draw, dashed, rounded corners, behind path}}
\tikzset{green-box/.style={box, fill=greenGood}}
\tikzset{blue-box/.style={box, fill=blue!20}}
\tikzset{bloom-filter/.style={
    draw,
    isosceles triangle,
    minimum width=1.2cm,
    shape border rotate=-90,
    isosceles triangle apex angle=60,
    rounded corners
}}
\tikzset{bw-delta/.style={draw, fill=redBad, rounded corners}}
\tikzset{bw-page/.style={draw, fill=greenGood, rounded corners}}
\tikzset{dp-tree-log/.style={draw, minimum width=0.5cm, minimum height=0.5cm, rounded corners=0.1cm}}
\tikzset{trie-node/.style={draw, circle}}

@@

    \def\arraystretch{1.5}
    \begin{tabular}{cccc}
        B-Tree & B$^{+}$-Tree & B$_{link}$-Tree & DP-Tree \\
        wB$^{+}$-Tree & NV-Tree & FP-Tree & FASTFAIR \\
        HiKV & Mass-Tree & Skip List & ART \\
        WORT & CDDS-Tree & Bw-Tree & HOT \\
        KISS-Tree & VAST-Tree & FAST & HV-Tree \\
        UB-Tree & LHAM & MDM & Hybrid B$^{+}$-Tree
    \end{tabular}
\end{frame}

@@

\end{frame}
\note{
    Postgres uses variation of B-link Tree, seems like MySQL innodb uses
    regular B+ Tree.
}

@@

\end{frame}

@@

    \end{center}
\end{frame}
\note{
    Another interesting approach is partial value on a page and the rest is on
    extended pages (MySQL, SQLite).
}

@@

    \begin{center}
    \textbf{Interpolation search}
    \vspace{1cm}

    \begin{overprint}[12cm]
        \onslide<1>
        \begin{center}
        \begin{tikzpicture}[]
            \node[draw] (key1) {5};
            \foreach \i / \value [evaluate=\i as \prev using \i-1] in
                {2/9,3/12,4/16,5/19,6/21,7/25,8/27,9/31,10/34,11/36}
            {
                \node[draw, right=0 of key\prev] (key\i) {\value};
            }
        \end{tikzpicture}
        \end{center}

        \onslide<2>
        \begin{center}
        \begin{tikzpicture}[]
            \node[draw] (key1) {5};
            \foreach \i / \value [evaluate=\i as \prev using \i-1] in
                {2/9,3/12,4/16,5/19,6/21,7/25,8/27,9/31,10/34,11/36}
            {
                \node[draw, right=0 of key\prev] (key\i) {\value};
            }

            \coordinate[below=2cm of key1] (start);
            \draw[->, btree-line] (start.south)
                .. controls ([xshift=1cm] start) and ([yshift=-3cm] key6) ..
                (key6.south);
            \draw[->, btree-line] (key6.south)
                .. controls ([yshift=-3cm] key6) and ([yshift=-3cm] key9) ..
                (key9.south);
            \draw[->, btree-line] (key9.south)
                .. controls ([yshift=-2cm] key9) and ([yshift=-2cm] key10) ..
                (key10.south);

        \onslide<3>
        \begin{center}
        \begin{tikzpicture}[xscale=0.6, yscale=0.1]
            \draw[->] (0,0) -- (11,0);
            \draw[->] (0,0) -- (0,36);
            \draw[color=red, line width=2pt] plot[id=x] file {\strongdata};
        \end{tikzpicture}
        \end{center}

        \onslide<4>
        \begin{center}
        \begin{tikzpicture}[]
            \node[draw] (key1) {5};
            \foreach \i / \value [evaluate=\i as \prev using \i-1] in
                {2/9,3/12,4/16,5/19,6/21,7/25,8/27,9/31,10/34,11/36}
            {
                \node[draw, right=0 of key\prev] (key\i) {\value};
            }

            \coordinate[below=2cm of key6] (start);
            \draw[->, btree-line] (start.south)
                .. controls ([xshift=1cm] start) and ([yshift=-3cm] key10) ..
                (key10.south);

        \onslide<5>
        \begin{center}
        \begin{tikzpicture}[]
            \node[draw] (key1) {5};
            \foreach \i / \value [evaluate=\i as \prev using \i-1] in
                {2/6,3/15,4/15,5/16,6/43,7/95,8/96,9/97,10/128,11/499}
            {
                \node[draw, right=0 of key\prev] (key\i) {\value};
            }

        \onslide<6>
        \begin{center}
        \begin{tikzpicture}[xscale=0.6, yscale=0.008]
            \draw[->] (0,0) -- (11,0);
            \draw[->] (0,0) -- (0,499);
            \draw[color=red, line width=2pt] plot[id=x] file {\weakdata};
        \end{tikzpicture}
        \end{center}
    \end{overprint}

@@

    \end{center}
\end{frame}
\note{
    Requires also "split", "index", "remove node", "merge", "flush" deltas
}

@@

@@

    \begin{center}
    \only<1> {
        \textbf{Is that all?}
    }
    \only<2> {
        \includegraphics[width=0.5\textwidth,center]{nein.png}
    }
    \only<3> {
    }

>>>>

\tikzset{rum-node/.style={draw,circle,fill=white,minimum width=2cm}}
\tikzset{rum-extra-node/.style={rectangle, draw, dashed, rounded corners, fill=white }}
\tikzset{btree-key/.style={minimum height=1cm, pattern=north west lines, draw}}
\tikzset{btree-pointer/.style={minimum height=1cm, draw}}
\tikzset{btree-hide/.style={draw opacity=0, line width=0, pattern=none}}
\tikzset{btree-empty/.style={btree-pointer, dashed}}
\tikzset{btree-leaf/.style={fill=greenGood}}
\tikzset{btree-branch/.style={fill=gray!20}}
\tikzset{btree-line/.style={line width=0.5pt}}
\tikzset{btree-path/.style={btree-pointer, minimum width=0.8cm, fill=red, opacity=0.5}}
\tikzset{>=latex}
\tikzset{btree-high-key/.style={btree-key, pattern=north west lines, pattern color=red, draw=red}}
\tikzset{btree-key-prefix/.style={btree-key, minimum height=0.2cm, fill=red, opacity=0.5}}
\tikzset{btree-key-compare/.style={btree-key, minimum height=0.6cm, fill=blue, opacity=0.5}}
\tikzset{arrow-pointer/.style={
        single arrow,
        minimum height=1.5cm,
        inner sep=3pt,
        line width=1pt,
        draw,
        color=gray,
        single arrow tip angle=45,
        single arrow head extend=0.1cm
    }
}
\tikzset{btree-indirection/.style={btree-key, minimum width=0.2cm, inner sep=0, pattern=none}}
\tikzset{btree-page-header/.style={btree-key, minimum width=0.5cm, inner sep=0, pattern=none}}
\tikzset{cpu-cache/.style={draw, minimum width=1cm, minimum height=1cm, text width=1cm, align=center, rounded corners}}
\tikzset{cpu-cache-hide/.style={cpu-cache, draw=none, color=white}}
\tikzset{btree-node/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners}}
\tikzset{btree-leaf-node/.style={btree-node, fill=greenGood}}
\tikzset{btree-branch-node/.style={btree-node, fill=gray!20}}
\tikzset{btree-partition1/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners, fill=red!20}}
\tikzset{btree-partition2/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners, fill=green!20}}
\tikzset{btree-partition3/.style={draw, inner sep=0, minimum width=1cm, minimum height=1cm, rounded corners, fill=blue!20}}
\tikzset{smallbtree-node/.style={btree-node, minimum width=0.4cm, minimum height=0.4cm, rounded corners=0.1cm}}
\tikzset{box/.style={draw, dashed, rounded corners, behind path}}
\tikzset{green-box/.style={box, fill=greenGood}}
\tikzset{blue-box/.style={box, fill=blue!20}}
\tikzset{bloom-filter/.style={
    draw,
    isosceles triangle,
    minimum width=1.2cm,
    shape border rotate=-90,
    isosceles triangle apex angle=60,
    rounded corners
}}
\tikzset{bw-delta/.style={draw, fill=redBad, rounded corners}}
\tikzset{bw-page/.style={draw, fill=greenGood, rounded corners}}
\tikzset{dp-tree-log/.style={draw, minimum width=0.5cm, minimum height=0.5cm, rounded corners=0.1cm}}
\tikzset{trie-node/.style={draw, circle}}
\tikzset{selected-path/.style={->, btree-line, line width=2pt, color=red}}

@@

    \def\arraystretch{1.5}
    \begin{tabular}{cccc}
        B-Tree & B$^{+}$-Tree & B$_{link}$-Tree & DPTree \\
        wB$^{+}$-Tree & NV-Tree & FPTree & FASTFAIR \\
        HiKV & Masstree & Skip List & ART \\
        WORT & CDDS-Tree & Bw-Tree & HOT \\
        KISS-Tree & VAST-Tree & FAST & HV-Tree \\
        UB-Tree & LHAM & MDAM & Hybrid B$^{+}$-Tree
    \end{tabular}
\end{frame}

@@

\end{frame}
\note{
    Postgres uses variation of B-link Tree, seems like MySQL innodb uses
    regular B+ Tree.

    WiredTiger uses parent pointers.
}

@@

\end{frame}
\note{
    B*-Tree rebalances between neighbors, e.g. as in SQLite.
}

@@

    \end{center}
\end{frame}
\note{
    Another interesting approach is partial value on a page and the rest is on
    extended pages (MySQL InnoDB, SQLite).
}

@@

    \begin{center}
    \textbf{Multidimentional Access Method}
    \vspace{1cm}

    \begin{overprint}[8cm]
        \onslide<1>
        \begin{tikzpicture}[]
            \basicbtree{show-connections}
            \node[below=0.2 of node5] {\footnotesize{1,a}};
            \node[below=1.8 of node2] {\footnotesize{...}};
            \node[below=0.2 of node6] {\footnotesize{1,z}};

            \node[below=0.2 of node7] {\footnotesize{2,a}};
            \node[below=1.8 of node3] {\footnotesize{...}};
            \node[below=0.2 of node8] {\footnotesize{2,z}};

            \node[below=0.2 of node9] {\footnotesize{3,a}};
            \node[below=1.8 of node4] {\footnotesize{...}};
            \node[below=0.2 of node10] {\footnotesize{3,z}};
        \end{tikzpicture}

        \onslide<2>
        \begin{tikzpicture}[]
            \basicbtree{show-connections}
            \node[below=0.2 of node5] {\footnotesize{1,a}};
            \node[below=1.8 of node2] {\footnotesize{...}};
            \node[below=0.2 of node6] {\footnotesize{1,z}};

            \node[below=0.2 of node7] {\footnotesize{2,a}};
            \node[below=1.8 of node3] {\footnotesize{...}};
            \node[below=0.2 of node8] {\footnotesize{2,z}};

            \node[below=0.2 of node9] {\footnotesize{3,a}};
            \node[below=1.8 of node4] {\footnotesize{...}};
            \node[below=0.2 of node10] {\footnotesize{3,z}};

            \draw[selected-path] (node1.center) -- (node2.center) -- (node5.center);
        \end{tikzpicture}

        \onslide<3>
        \begin{tikzpicture}[]
            \basicbtree{show-connections}
            \node[below=0.2 of node5] {\footnotesize{1,a}};
            \node[below=1.8 of node2] {\footnotesize{...}};
            \node[below=0.2 of node6] {\footnotesize{1,z}};

            \node[below=0.2 of node7] {\footnotesize{2,a}};
            \node[below=1.8 of node3] {\footnotesize{...}};
            \node[below=0.2 of node8] {\footnotesize{2,z}};

            \node[below=0.2 of node9] {\footnotesize{3,a}};
            \node[below=1.8 of node4] {\footnotesize{...}};
            \node[below=0.2 of node10] {\footnotesize{3,z}};

            \draw[selected-path] (node1.center) -- (node3.center) -- (node7.center);
        \end{tikzpicture}

        \onslide<4>
        \begin{tikzpicture}[]
            \basicbtree{show-connections}
            \node[below=0.2 of node5] {\footnotesize{1,a}};
            \node[below=1.8 of node2] {\footnotesize{...}};
            \node[below=0.2 of node6] {\footnotesize{1,z}};

            \node[below=0.2 of node7] {\footnotesize{2,a}};
            \node[below=1.8 of node3] {\footnotesize{...}};
            \node[below=0.2 of node8] {\footnotesize{2,z}};

            \node[below=0.2 of node9] {\footnotesize{3,a}};
            \node[below=1.8 of node4] {\footnotesize{...}};
            \node[below=0.2 of node10] {\footnotesize{3,z}};

            \draw[selected-path] (node1.center) -- (node4.center) -- (node9.center);
        \end{tikzpicture}
    \end{overprint}

    \end{center}
\end{frame}
\note{
    For not complete search keys, if a dimension is under-specified, its
    possible values must be enumerated.
}

%\begin{frame}[fragile]{}
    %\frametitle{}

    %\begin{center}
    %\textbf{Interpolation search}
    %\vspace{1cm}

    %\begin{overprint}[12cm]
        %\onslide<1>
        %\begin{center}
        %\begin{tikzpicture}[]
            %\node[draw] (key1) {5};
            %\foreach \i / \value [evaluate=\i as \prev using \i-1] in
                %{2/9,3/12,4/16,5/19,6/21,7/25,8/27,9/31,10/34,11/36}
            %{
                %\node[draw, right=0 of key\prev] (key\i) {\value};
            %}
        %\end{tikzpicture}
        %\end{center}

        %\onslide<2>
        %\begin{center}
        %\begin{tikzpicture}[]
            %\node[draw] (key1) {5};
            %\foreach \i / \value [evaluate=\i as \prev using \i-1] in
                %{2/9,3/12,4/16,5/19,6/21,7/25,8/27,9/31,10/34,11/36}
            %{
                %\node[draw, right=0 of key\prev] (key\i) {\value};
            %}

            %\coordinate[below=2cm of key1] (start);
            %\draw[->, btree-line] (start.south)
                %.. controls ([xshift=1cm] start) and ([yshift=-3cm] key6) ..
                %(key6.south);
            %\draw[->, btree-line] (key6.south)
                %.. controls ([yshift=-3cm] key6) and ([yshift=-3cm] key9) ..
                %(key9.south);
            %\draw[->, btree-line] (key9.south)
                %.. controls ([yshift=-2cm] key9) and ([yshift=-2cm] key10) ..
                %(key10.south);

        %\end{tikzpicture}
        %\end{center}

        %\onslide<3>
        %\begin{center}
        %\begin{tikzpicture}[xscale=0.6, yscale=0.1]
            %\draw[->] (0,0) -- (11,0);
            %\draw[->] (0,0) -- (0,36);
            %\draw[color=red, line width=2pt] plot[id=x] file {\strongdata};
        %\end{tikzpicture}
        %\end{center}

        %\onslide<4>
        %\begin{center}
        %\begin{tikzpicture}[]
            %\node[draw] (key1) {5};
            %\foreach \i / \value [evaluate=\i as \prev using \i-1] in
                %{2/9,3/12,4/16,5/19,6/21,7/25,8/27,9/31,10/34,11/36}
            %{
                %\node[draw, right=0 of key\prev] (key\i) {\value};
            %}

            %\coordinate[below=2cm of key6] (start);
            %\draw[->, btree-line] (start.south)
                %.. controls ([xshift=1cm] start) and ([yshift=-3cm] key10) ..
                %(key10.south);

        %\end{tikzpicture}
        %\end{center}

        %\onslide<5>
        %\begin{center}
        %\begin{tikzpicture}[]
            %\node[draw] (key1) {5};
            %\foreach \i / \value [evaluate=\i as \prev using \i-1] in
                %{2/6,3/15,4/15,5/16,6/43,7/95,8/96,9/97,10/128,11/499}
            %{
                %\node[draw, right=0 of key\prev] (key\i) {\value};
            %}

        %\end{tikzpicture}
        %\end{center}

        %\onslide<6>
        %\begin{center}
        %\begin{tikzpicture}[xscale=0.6, yscale=0.008]
            %\draw[->] (0,0) -- (11,0);
            %\draw[->] (0,0) -- (0,499);
            %\draw[color=red, line width=2pt] plot[id=x] file {\weakdata};
        %\end{tikzpicture}
        %\end{center}
    %\end{overprint}

    %\end{center}
%\end{frame}

@@

    \end{center}
\end{frame}
\note{
    Requires also "split", "index", "remove node", "merge", "flush" deltas
    MSFT Hekaton, Sled db - Bw-Tree
}

@@

    \end{center}
\end{frame}
\note{
    Silo - Masstree
    HyPer - ART
    ClickHouse - BTrie
}

@@

    \begin{center}
    \only<1> {
        \textbf{Is that all?}
    }
    \only<2> {
        \includegraphics[width=0.5\textwidth,center]{nein.png}
        \vspace{-0.8cm}
        \hspace*{6cm}
        \href{https://www.youtube.com/watch?v=_dRra4786Ww}
             {\color{white}\fontsize{5pt}{0}\selectfont @Unzucht}

    }
    \end{center}
\end{frame}

    \begin{center}
        \begin{itemize}[label={\MVRightarrow}]
            \item <+-> Buffered (Lazy) B-Tree
            \item <+-> Cache conscious B-Tree
            \item <+-> Learned indexes
            \item <+-> UB-Tree
            \item <+-> Skip List
            \item <+-> LSM Tree
            \item <+-> ...
        \end{itemize}
    \end{center}
\end{frame}
\note{
    Buffered B-Tree are used in WiredTiger.
    Skip list are used also in WiredTiger as in memory structure, in Apache
    Cassandra again as in memory, and in Redis.
}
